---
title: "Juvenile Snowshoe hare physical data: An exploratory report"
author: "Erin de Leon Sanchez"
date: "11/7/2021"
output:
html_document: 
    theme: spacelab

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(kableExtra)
library(effsize)
library(broom)
library(equatiomatic)
```

### A. Introduction: 
This report covers an exploratory overview of juvenile Snowshoe hare physical data from the Bonanza Creek Long Term Ecological Research (LTER) site from 1999 - 2012. The Snowshoe hare, *Lepus americanus*, is a keystone prey species in northern boreal forests and experience population fluctuations of 8-11-years and these data were collected to estimate Snowshoe hare population densities and growth over time. Here, annual juvenile hare trap counts are reported to see population size fluctuation over time. To explore the physical data, we report a comparison of juvenile hare weights by sex and site (grid) as well as a juvenile weight comparison by sex. We conclude the report with an exploration of the relationship between juvenile Snowshoe hare weight and hind foot length. 

### B. Data and analyses:
Juvenile snowshoe hare physical data were collected and made available by Knut Kielland, F. Stuart Chapin, and Roger W. Ruess at the Bonanza Creek Experimental Forest LTER, approximately 20 km southwest of Fairbanks, Alaska in the Tanana Valley State Forest. These data are from capture-recapture studies of snowshoe hares at 5 locales, although only three sites (grids) are included in this report, in the Tanana valley from 1999 to 2012. Following exploratory data visualization, weights for male and female juvenile hares are compared through Effect size (Cohen's *d*) an unpaired two-sample t-test. The relationship between weight and hind foot length is explored by linear regression. All analyses are in R version 3.6.2 using RStudio version RStudio 2021.09.0+351 for macOS.

```{r}
#Read in the data 
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))
```

#### i. Annual juvenile Snowshoe hare trap counts 
```{r}
# First convert the date to class “Date” so that I can create a new column that only contains the year
total_trap <- bonanza_hares %>% 
  mutate(site_full = case_when(
    grid == "bonbs"  ~ "Bonanza Riparian", 
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Black Spruce")) %>% # Added a new column with the full site name
  mutate(sex_full = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male")) %>% # Added a new column with each sex fully spelled out
  mutate(date_new = as_date(date, format="%m/%d/%y")) %>% # First convert the date to class “Date” 
  mutate(year = year(date_new)) %>% # Create a new column for year
  filter(age %in% "j") %>% # Filtered for juvenile hares
  group_by(year) %>% # Group by year
  summarize(sample_size = n()) # Count the total number of trappings for each year
```
```{r}
# Data visualization
ggplot(data = total_trap, aes(x = year, y = sample_size)) +
  geom_col() +
  scale_x_continuous(breaks = 1999:2012) +
  scale_y_continuous(limits=c(0, 130), breaks = c(0, 25, 50, 75, 100, 125)) +
  theme_light() +
  labs(x = "Year", y = "Total number of juvenile hare trappings", title = "Total number of juvenile hare trappings from 1999 to 2012", caption = "Data from Bonanza Creek Long Term Ecological Research")
```

**Figure 1:** The number of juvenile hare trap counts per year from 1999 to 2012. Note that these are absolute counts.
The minimum count of hare trappings is 2 and the maximum count is 126 hare trappings. The mean annual number of juvenile hares trapped is 31.5 and the median number of juvenile hares trapped is 18.5, indicating that these counts are not normally distributed among the years of the study. To standardize the juvenile hare population in future analyses... 

```{r, include = FALSE}
# Some summary stats to comment on 
summary_hares <- total_trap %>% 
  summarize(mean_hares = mean(sample_size, na.rm = TRUE), 
            median_hares = median(sample_size, na.rm = TRUE),
            sd_hares = sd(sample_size, na.rm = TRUE), 
            min(sample_size),
            max(sample_size))
summary_hares
```


#### ii. A comparison of juvenile Snowshoe hare weights by sex and site (grid)

```{r}
# Preparing data for visualization
weight_sex_grid <- bonanza_hares %>% # created a new df
  mutate(site_full = case_when( 
    grid == "bonbs"  ~ "Bonanza Riparian", 
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Black Spruce")) %>% # Added a new column with full site name
  mutate(Sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male")) %>% # Added a new column with sex completely spelled out
  mutate(date_new = as_date(date, format="%m/%d/%y")) %>% # First converted the date to class “Date” 
  mutate(year = year(date_new)) %>% # Created a new column for year
  filter(age %in% "j") %>% # Filtered for juvenile hares
  select(weight, site_full, Sex, age) %>% # Kept columns that I needed 
  group_by(site_full) 
```

```{r}
# Data as a histogram
ggplot(data = weight_sex_grid, aes(x = weight, fill = Sex, color = Sex)) +
  geom_histogram(alpha = 0.5) +
  facet_wrap(~ site_full) +
  theme_bw() +
  labs(x = "Weight (g)", y = "Number of hares", title = "Juvenile hare weights by sex and site", caption = "Data from Bonanza Creek Long Term Ecological Research")
  

#ggplot(data = weight_sex_grid, aes(x = Sex, y = weight, color = Sex)) +
  #geom_boxplot() +
  #facet_wrap(~ site_full) +
  #theme_bw() 

```

**Figure 2:** Juvenile Snowshoe hare weight distribution by sex and site. 

Main takeaways:

- Weight distribution of both juvenile male and female hares is not normally distributed, regardless of site.
- There are greater numbers of juvenile female hares distributed across the range of weights across the three sites, as compared to juvenile male hares.


#### iii. Juvenile Snowshoe hare weight comparison between sexes
```{r, include=FALSE}

# Pull vectors of weight for male and female juvenile hares
m_hares <- bonanza_hares %>% 
  filter(sex == "m", age == "j") %>% 
  filter(!is.na(weight)) %>% 
  pull(weight)

f_hares <- bonanza_hares %>% 
  filter(sex == "f", age == "j") %>% 
  filter(!is.na(weight)) %>% 
  pull(weight)

# Plot histogram and qq plots to check assumptions (for t-test)
hist(m_hares) 
qqnorm(m_hares) 
hist(f_hares) 
qqnorm(f_hares)

# Run the t-test:
hares_ttest <- t.test(m_hares, f_hares) 
hares_ttest

# Get the tidy model results to call outputs in-line: 
hares_ttest_tidy <- tidy(hares_ttest)
hares_ttest_tidy

# Get the means & SDs for each: 
male_mean <- mean(m_hares)
female_mean <- mean(f_hares)
male_sd <- sd(m_hares)
female_sd <- sd(f_hares)


# Get the effect size: 
hare_effsize <- cohen.d(m_hares, f_hares) 
hare_effsize
```
On average, juvenile male Snowshoe hares weigh more than juvenile female Snowshoe hares (`r round(male_mean,2)` $\pm$ `r round(male_sd, 2)` and `r round(female_mean,2)` $\pm$ `r round(female_sd, 2)` grams, respectively; mean $\pm$ 1 standard deviation). While the absolute difference in means is `r round(male_mean, 2) - round(female_mean, 2)` mm (a `r round((male_mean - female_mean)/((male_mean + female_mean)/2)*100, 2)`% difference), the difference in means is significant (Welch's two-sample t-test: t(`r round(hares_ttest_tidy$parameter,2)`) = `r round(hares_ttest_tidy$statistic,2)`, p < 0.05), and the effect size is small (Cohen's *d* = `r round(hare_effsize$estimate,2)`).

```{r, include = TRUE}
summary_hares <- weight_sex_grid %>% 
  group_by(Sex) %>% 
  select(Sex,weight) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            sd_weight = sd(weight, na.rm = TRUE), 
            sample_size = n())

kable(summary_hares, col.names = c("Sex", "Mean weight (g)", "Standard Deviation (g)", "Sample size (n)"), digits = 2, caption = "Table 1: Summary statistics for juvenile snowshoe hares") %>% 
kable_classic()
```



#### iv. Relationship between juvenile Snowshoe hare weight & hind foot length.

```{r}
hindft_weight <- bonanza_hares %>% 
  filter(age %in% "j") %>% 
  select(weight, hindft) 

ggplot(data = hindft_weight, aes(x = hindft, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Hind foot length (mm)", y = "Weight (g)", title = "Juvenile hare hind foot length vs. weight", caption = "Data from Bonanza Creek Long Term Ecological Research") +
  theme_light()

```

**Figure 2:** 
In 2 - 3 sentences below your figure, describe the outcome of your linear regression exploration (e.g. slope interpretation, R2 value, and Pearson's r correlation) and touch briefly on which assumption(s) of linear regression may be a concern, and additional thoughts you have about how well (or not well…) the linear model describes the relationship. What worries you? 



```{r, include = FALSE}
#linear regression
hindft_weight_lm <- lm(weight ~ hindft, data = hindft_weight) 

# Get tidy versions of the model output to call later on in text
hindft_weight_lm_tidy <- broom::tidy(hindft_weight_lm)
broom::glance(hindft_weight_lm) 

#See model diagnostics
plot(hindft_weight_lm)

# Get Pearson's r correlation
hares_cor <- cor.test(hindft_weight$weight, hindft_weight$hindft)

# Tidy version of correlation output:
hares_cor_tidy <- tidy(hares_cor)
```


### D. Summary: 
Exploratory data analysis reveals the following initial findings:

- swag
- cool
- yasss

### Citations: 

Data citation: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

